Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2022-11-15 reports/supplement/supplement_0.9.1.Rmd
update supplement

```{r echo = FALSE}
rm(list = ls())
library(tidyverse)
library(microxanox)
library(patchwork)
library(here)
source(here("r", "various_useful_functions.r"))
# source(here("experiments", "1_figures_main_ms", "ms_figure_functions.r"))
source(here("R/ms_figure_functions.R"))

@@

As is shown in the figures below, the temporal method results in a larger region of bistability than the replication method, but only during increasing oxygen diffusivities. The difference in results occurs because, as discussed below, the initial conditions differ between the two approaches.

@@

The replication method shows the same effects of functional diversity as those revealed by the temporal method (Figure \@ref(fig:div-repl-meth1)). Namely

-   Introducing functional diversity in only the cyanobacteria increases the resilience of the oxic state but has no effect on the resilience of the anoxic state. 
-   Introducing functional diversity in only the sulfate-reducing bacteria increases the resilience of the anoxic state and reduces the resilience of the oxic state.
-   Introducing functional diversity in only the phototrophic sulfur bacteria reduces the resilience of the anoxic state and has no effect on the resilience of the oxic state. 
-   Introducing functional diversity in more than one functional group reduces, reverses, or does not change the effect of diversity in individual functional groups. For example, introducing functional diversity in all three functional groups reduces the resilience of the oxic state when trait variation is low but increases the resilience of the oxic state when trait variation is high.

@@

Figure \@ref(fig:repl-meth-ex1) shows an example of stable states along the oxygen diffusivity gradient produced by the replication method for finding stable states. The patterns are similar to those produced by the temporal method for finding the stable states (see Figure 4 in the main report), though they differ in one respect: close to the transition back to the state a group dominates, some of the more tolerant strains occur in the stable states found by the temporal method (Figure 4 in the main report), while they do not occur in stable states found by the replication method (Figure \@ref(fig:repl-meth-ex1) here). We understand that the appearance of these strains is transient, and the difference in their appearance results from the differences in starting conditions of the two methods. Specifically, in the temporal method, the initial condition at a new level of oxygen diffusivity is the stable state at the previous level of oxygen diffusivity, while in the replication method the starting conditions are identical at each level of oxygen diffusivity. These differences in starting conditions apparently result in longer persistence of transients in the temporal method than the replication method.

@@

```{r div-temp-eventdef1, eval = FALSE, echo = FALSE,  include=TRUE, fig.cap='(ref:div-temp-eventdef1)', fig.height=8, fig.width=8, fig.align="center"}
all_stab <- readRDS(here("data",
                          microxanox_version_path,
                          "temporal_method",
                          "event_definition_1",
                          "processed_data",
                         "all_stab_data.RDS"))

sim_length <- 1e6
num_strains <- 9
plot_here <- all_stab %>%
  filter(waittime == wait_time)
p1 <- fig_div_vs_o2diff_1strain_7row(plot_here,
                                     which_strain = num_strains,
                                     figure_title = paste0(sim_length,
                                                           " simulation length,\n",
                                                           num_strains,
                                                           " strains"))
p2 <- fig_resilience_vs_div(plot_here,
                            which_strain = num_strains,
                            figure_title = " ")
p1 + p2

@@

A particularly notable difference between the two approaches is in the initial condition. In the temporal approach, the initial condition at a new level of oxygen diffusivity is the stable state at the previous level of oxygen diffusivity, while in the approach of [Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x) the initial conditions are fixed and arbitrary (though this approach is still useful for illustrating the bistability of the system). This difference in initial conditions increases the region of bistability in the temporal approach because as a tipping point approaches, the functional group that will dominate after the regime shift starts from much lower abundances than in the approach by [Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x) and thus requires stronger amelioration of the environment to take over. Nevertheless, our main conclusions about the effects of functional trait diversity are the same using either approach.

@@

The results below show the stable states as a function of oxygen diffusivity for a system with an intermediate level of diversity in all functional groups, with either 1'000'000 hours at each oxygen diffusivity level (Figure \@ref(fig:ss-temp-meth3)), or 2'000'000 hours (Figure \@ref(fig:ss-temp-meth4)). This shows that 1'000'000 hours duration of the simulations are sufficient to reach stable state, except very close to state transitions on the trajectory of recovery (i.e.: cyanobacteria: trajectory of increasing oxygen diffusivity; sulfate-reducing bacteria: trajectory of decreasing oxygen diffusivit). Very close to these transitions, even at 2'000'000 hours, the dynamics are still transient, with high tolerance strains at high density, even though they will eventually be outcompeted by low tolerance (higher growth rate) strains.

@@

```{r ss-temp-meth3, eval = FALSE, echo = FALSE,  include=TRUE, fig.cap='(ref:ss-temp-meth3)', fig.height=10, fig.width=8, fig.align="center"}

## data made with the "run_single_sim.r" file in the ss_find, temporal method folder.

@@

```{r ss-temp-meth4, eval = FALSE, echo = FALSE,  include=TRUE, fig.cap='(ref:ss-temp-meth4)', fig.height=10, fig.width=8, fig.align="center"}
single_2e6 <- readRDS(here("data",
                           microxanox_version_path, 
                           "temporal_method",
                           "single_sims",
                           "single_2e6_result.RDS"))
temp_meth[1,]$ssfind_result[[1]] <- single_2e6
fig_state_vs_o2diff_sidebyside_alternative(temp_meth[1,])
```

@@

# Are the results driven by functional diversity?

During the review process we received the following comment: *I think the authors have not shown sufficiently clearly that their results are actually driven by functional diversity. Fig. 4 shows that on the collapse trajectory, the dominating functional group is itself dominated by the most tolerant strain. This is also true for the suppressed group on its recovery trajectory shortly before the tipping point. This made me wonder whether the same (or at least a very similar) results regarding the change in resilience of the system states could also be obtained with just a single strain per functional group, whose tolerance to the inhibiting substance is systematically increased. This could be tested fairly easily by re-running the simulations behind Fig. 3 with just the most tolerant (and slowest growing) strain of each functional group present.* In this section we discuss this matter.

Our first realisation was that we had not adequately stated what component of functional diversity we manipulated. We revised to state *Functional diversity has multiple components, and we manipulate only one of these, namely the range of trait space represented, which has been termed functional richness, and measured as the convex hull volume ([Villéger et al. 2008](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/07-1206.1)).*

Since we manipulated the range of trait space represented (sensu [Villéger et al 2008](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/07-1206.1)), and manipulated nothing else, we manipulated only functional diversity. Logically then, any effects of this manipulation must be due to the functional diversity manipulation. That is, regardless of the process / mechanism by which functional diversity causes an effect, it is appropriate to state that functional diversity drives the results.

The reviewer’s comment motivated us to consider two questions:

1. By what mechanism does functional diversity affect ecosystem stability in this study? 
2. Could the observed effects of functional diversity also be created by manipulations of trait values that do not alter functional diversity?

Question 1. In research concerning biodiversity and ecosystem functioning there was considerable debate about the phenomenon that high yielding species are more likely to be present in more diverse communities. This phenomenon was sometimes responsible for a large proportion of observed effects of biodiversity on ecosystem function. The debate was to a large extent resolved by recognising that biodiversity effects can result from two sources: a selection/sampling effect, and a complementarity effect. Both are now considered valid and important mechanisms by which diversity (including functional diversity) affects ecosystem functioning. We view the presence of more extreme trait values in more functionally diverse communities as an analogue of the sampling/selection effect of biodiversity, and a valid mechanism by which increased functional diversity can affect ecosystem stability. For example, let us assume for the moment that, indeed, only the single maximum tolerance strain present in each functional group of an ecosystem determines the resilience of that ecosystem. Then we would say that variation in resilience with diversity operates only by the "sampling effect". If we also assume that higher tolerance causes higher ecosystem resilience, then a more functionally diverse ecosystem will have higher maximum tolerance trait, and so higher resilience. And an ecosystem with more strains per functional group will have higher resilience (Figure \@ref(fig:samp-effect1)).

Hence, even if all of the observed effects of our functional diversity manipulation were only due to presence of one strain (e.g. the most tolerant one), this can be considered an effect of functional diversity.

@@

For most of the explored combinations and amounts of functional diversity/tolerance the tipping point positions did not differ between the diversity manipulation and the tolerance manipulation (Figure \@ref(fig:max-tol1)). However, the tipping point positions did differ between the diversity manipulation and the tolerance manipulation when diversity/tolerance in the SB group was high (i.e. in the combinations SB only, CB-SB, and CB-SB-PB). We show below that

(ref:max-tol1) Effects of strain variation on position of tipping points compared between simulations with all (nine) strains present and simulations where only the most tolerant strain is present. In the SB and CB-SB treatments there is missing data in the case with only the most tolerant strain when there were no alternate stable states and therefore no tipping points.

@@

## remove data where there are no stable states
## this subset is identified by inspection of the figure
all_stab <- all_stab %>%
  filter(!(var_treat == "SB" & stand_var > 0.88 & num_strains == -1),
         !(var_treat == "CB-SB" & stand_var > 0.84 & num_strains == -1))

@@

When the SB contained only one strain with very high tolerance and low maximum growth rate, there were no alternate stable states, i.e. the state of the system was independent of the direction of change in oxygen diffusivity (Figure \@ref(fig:max-tol3)). Furthermore, at low oxygen diffusivity the system varied around a stable state with all three functional groups coexisting –this never occurred when there were multiple strains varying in tolerance (Figure \@ref(fig:max-tol2)), or when there was only one strain with mean tolerance.

Apparently, absence of SB strains with high growth rate precludes the formation of a stable state that does not include cyanobacteria. It seems that the most tolerant SB strain does not have a high enough growth rate to itself exclude the cyanobacteria. Cyanobacteria are therefore always present, regardless of the oxygen diffusivity.

@@

```{r max-tol3, echo = FALSE,  include=TRUE, fig.cap='(ref:max-tol3)', fig.height=10, fig.width=8, fig.align="center"}
max_tol_one <- filter(maxtolonly_strains,
       CB_var_gmax_s == 0,
       abs(SB_var_gmax_s - 0.09473684) < 0.001,
       PB_var_gmax_s == 0)
fig_state_vs_o2diff_sidebyside_alternative_nofliparrows(max_tol_one)
```

### What causes the variation in response to oxygen diffusivity (SB)?

@@

## High tolerance vs. high diversity in CB and SB

As in the SB only case, the absence of lower tolerance-higher growth rate strains changes the system to a parameter space where it does not display alternate states, at least within the range of oxygen diffusivities simulated (Figure \@ref(fig:max-tol5)).

(ref:max-tol4)  Stables states found when there is high trait variation among nine strains of CB and SB. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

```{r max-tol4, echo = FALSE,  include=TRUE, fig.cap='(ref:max-tol4)', fig.height=10, fig.width=8, fig.align="center"}
filter(nine_strains,
       abs(CB_var_gmax_s - 0.026592798) < 0.001,
       abs(SB_var_gmax_s - 0.079778395) < 0.001,
       PB_var_gmax_s == 0) %>%
fig_state_vs_o2diff_sidebyside_alternative_leftfliponly()
```

@@

```{r max-tol5, echo = FALSE,  include=TRUE, fig.cap='(ref:max-tol5)', fig.height=10, fig.width=8, fig.align="center"}
max_tol_one_CBSB <- filter(maxtolonly_strains,
       abs(CB_var_gmax_s - 0.026592798) < 0.001,
       abs(SB_var_gmax_s - 0.079778395) < 0.001,
       PB_var_gmax_s == 0)
fig_state_vs_o2diff_sidebyside_alternative_nofliparrows(max_tol_one_CBSB)
```

### What causes the variation in response to oxygen diffusivity (CB-SB)?

In the ecosystem shown in Figure \@ref(fig:max-tol5), where CB and SB each contained only one strain with very high tolerance, there was large variation in state variables along the oxygen diffusivity gradient at values less than about -6.5. To examine what might be causing this, we simulated the ecosystem at a constant oxygen diffusivity of $log_{10}$ oxygen diffusivity of -7 (Figure \@ref(fig:CBSB-maxtol-dyn1)). We see that the dynamics are qualitatively different from other cases in which there is a stable state reached in such a simulation (Figure \@ref(fig:sb-div-effect4)). It appears that the SB strain with high tolerance is able to initiate a transition to an anoxic state, but is not able to create a stable anoxic state.

```{r echo = FALSE}
#nine_strain_one$pars[[1]]
#max_tol_one$pars[[1]]
#fig_state_vs_o2diff_sidebyside_alternative(nine_strain_one)
#fig_state_vs_o2diff_sidebyside_alternative(max_tol_one)
parameter <- max_tol_one_CBSB$parameter[[1]]
parameter$strain_parameter <- max_tol_one$pars[[1]]
#parameter$strain_parameter <- ss_result$pars[[1]]
total_initial_abundances <- 10^5
parameter <- set_temporal_ssfind_initial_state(parameter,
                                               total_initial_abundances,
                                               total_initial_abundances,
                                               total_initial_abundances)
parameter$log10a_series <- c(-7, -7)
wait_time <-  1e5
parameter$sim_duration <- wait_time * length(parameter$log10a_series)
parameter$sim_sample_interval <- 10
#parameter$sim_duration
test1 <- run_simulation(parameter)
```

(ref:CBSB-maxtol-dyn1) Temporal dynamics of an ecosystem with a single CB strain with high tolerance and a single SB strain with high tolerance, the same one as in Figure \@ref(fig:max-tol5) simulated with constant $log_{10}(oxygen diffusivity)$ of -6.

```{r CBSB-maxtol-dyn1, echo = FALSE,  include=TRUE, fig.cap='(ref:CBSB-maxtol-dyn1)', fig.height=10, fig.width=8, fig.align="center"}
p <- plot_dynamics(test1)
p[[1]] <- p[[1]] + 
  xlab(NULL) + 
  ylab(expression(log[10](density)~(cells~L^{-1}))) +
  theme_bw()
p[[2]] <- p[[2]] + 
  xlab(NULL) + 
  ylab(expression(log[10](density)~(cells~L^{-1}))) +
  theme_bw()
p[[3]] <- p[[3]] + 
  xlab(NULL) + 
  ylab(expression(log[10](density)~(cells~L^{-1}))) +
  theme_bw()
p[[4]]$data <- p[[4]]$data %>% 
  dplyr::filter(species != "a")
p[[4]]$data$species <- p[[4]]$data$species %>%
  dplyr::recode(SO = "Oxidised sulfur", O = "Oxygen", SR = "Sulfide", P = "Phosphorus")
p[[4]] <- p[[4]] + 
  xlab("time (h)") + 
  ylab(expression(log[10](concentration)~(µM))) + 
  labs(colour = "substrate") +
  theme_bw()

@@

When in each of the three functional groups only the most tolerant strain is present, the resilience of the oxic state is greater than when there is variation in strain tolerance within groups. Looked at from another perspective, comparing CB-SB and CB-SB-PB both with only the most tolerant strains, presence of a high tolerance PB strain (in CB-SB-PB) allows the formation of an anoxic cyanobacteria-free stable state (whereas there was no such state in CB-SB). We interpret this as presence of a high tolerance PB strain stabilising the anoxic state, perhaps by increasing the availability of oxidised sulfur which is a resource for SB. Nevertheless, absence of SB strains with high growth rate seems to shift the oxic-anoxic tipping point to lower oxygen diffusivity.

@@

When two functional groups varied in traits, the effects of trait variation were additive when one of the groups had no diversity effect on a tipping point but non-additive when both groups had diversity effects on a tipping point (Fig. \@ref(fig:fig-add-linear) a-f). High trait variation in cyanobacteria and sulfate-reducing bacteria caused the oxic-anoxic tipping point to occur at lower oxygen diffusivity than expected assuming additivity of the individual effects of diversity in these two groups (Fig. \@ref(fig:fig-add-linear)d). When the two groups of sulfur bacteria varied in traits, the anoxic-oxic tipping point occurred at lower oxygen diffusivity than expected by additivity (Fig. \@ref(fig:fig-add-linear)f).

When all three groups varied in traits, only the diversity effect in phototrophic sulfur bacteria on the oxic-anoxic tipping point was additive; all other effects were non-additive, especially at higher trait variation (Fig. \@ref(fig:fig-add-linear)g-i). Both tipping points occurred at lower levels of oxygen diffusivity than expected by additivity. Interestingly, diversity in the cyanobacteria moved the shift to the oxic state to lower levels of oxygen diffusivity although independent variation in the cyanobacteria had no effect on this tipping point.

>>>>

```{r echo = FALSE}
rm(list = ls())
library(tidyverse)
library(microxanox)
library(patchwork)
library(here)
source(here("r", "various_useful_functions.R"))
# source(here("experiments", "1_figures_main_ms", "ms_figure_functions.R"))
source(here("R/ms_figure_functions.R"))

@@

As is shown in the figures below, the temporal and replication methods result in very similar regions of bistability.

@@

The replication method shows qualitatively the same effects of functional diversity (Figure \@ref(fig:div-repl-meth1)) as those revealed by the temporal method. Namely

-   Introducing functional diversity in only the cyanobacteria increases the resilience of the oxic state and slightly decreases the resilience of the anoxic state. 
-   Introducing functional diversity in only the sulfate-reducing bacteria increases the resilience of the anoxic state and reduces the resilience of the oxic state.
-   Introducing functional diversity in only the phototrophic sulfur bacteria reduces the resilience of the anoxic state and has no effect on the resilience of the oxic state. 
-   Introducing functional diversity in more than one functional group reduces, reverses, or does not change the effect of diversity in individual functional groups. For example, introducing functional diversity in all three functional groups reduces the resilience of the oxic state when trait variation is low but increases the resilience of the oxic state when trait variation is high.

@@

Figure \@ref(fig:repl-meth-ex1) shows an example of stable states along the oxygen diffusivity gradient produced by the replication method for finding stable states. The patterns are similar to those produced by the temporal method for finding the stable states (see Figure 4 in the main report), though they differ in one respect: close to the transition back to the state a group dominates, some of the more tolerant strains occur in the stable states found by the temporal method (Figure 4 in the main report), while they do not occur in stable states found by the replication method (Figure \@ref(fig:repl-meth-ex1) here). We understand that the appearance of these strains is transient, and the difference in their appearance results from the differences in starting conditions of the two methods. Specifically, in the temporal method, the initial condition at a new level of oxygen diffusivity is the stable state at the previous level of oxygen diffusivity, while in the replication method the starting conditions are identical at each level of oxygen diffusivity. These differences in starting conditions apparently result in longer duration of transients in the temporal method than the replication method.

@@

```{r div-temp-eventdef1, eval = TRUE, echo = FALSE,  include=TRUE, fig.cap='(ref:div-temp-eventdef1)', fig.height=8, fig.width=8, fig.align="center"}
all_stab <- readRDS(here("data",
                          microxanox_version_path,
                          "temporal_method",
                          "event_definition_1",
                          "processed_data",
                         "all_stab_data.RDS"))

@@

A particularly notable difference between the two approaches is in the initial condition. In the temporal approach, the initial condition at a new level of oxygen diffusivity is the stable state at the previous level of oxygen diffusivity, while in the approach of [Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x) the initial conditions are fixed and arbitrary (though this approach is still useful for illustrating the bistability of the system).

@@

The results below show the stable states as a function of oxygen diffusivity for a system with an intermediate level of diversity in all functional groups, with either 1'000'000 hours at each oxygen diffusivity level (Figure \@ref(fig:ss-temp-meth3)), or 2'000'000 hours (Figure \@ref(fig:ss-temp-meth4)). This shows that 1'000'000 hours duration of the simulations are sufficient to reach stable state, except very close to state transitions on the trajectory of recovery (i.e.: cyanobacteria: trajectory of increasing oxygen diffusivity; sulfate-reducing bacteria: trajectory of decreasing oxygen diffusivity). Very close to these transitions, even at 2'000'000 hours, the dynamics are still transient, with high tolerance strains at high density, even though they will eventually be outcompeted by low tolerance (higher growth rate) strains.

@@

```{r ss-temp-meth3, eval = TRUE, echo = FALSE,  include=TRUE, fig.cap='(ref:ss-temp-meth3)', fig.height=10, fig.width=8, fig.align="center"}

temp_meth <- readRDS(here("data",
                          microxanox_version_path,
                          "temporal_method",
                          "event_definition_2",
                          "ss_data_9strains_waittime1e+06_event_definition_2.RDS"))

## data made with the "run_single_sim.R" file in the ss_find, temporal method folder.

@@

```{r ss-temp-meth4, eval = TRUE, echo = FALSE,  include=TRUE, fig.cap='(ref:ss-temp-meth4)', fig.height=10, fig.width=8, fig.align="center"}
single_2e6 <- readRDS(here("data",
                           microxanox_version_path, 
                           "temporal_method",
                           "single_sims",
                           "single_2e6_result.RDS"))
temp_meth[1,]$ssfind_result[[1]] <- single_2e6
fig_state_vs_o2diff_sidebyside_alternative(temp_meth[1,])
```

@@

# Underlying mechanisms

In the main text we show that trait variation can affect resilience. In this supplementary section, we discuss the mechanism(s) underlying the observed diversity effects. We first explain the two major classes of mechanisms known to drive diversity effects, and then use additional simulations to elucidate the mechanisms operating in our study.

## Complementarity and sampling effect

Research about the relationship between biodiversity and ecosystem functioning has identified two major mechanisms by which diversity affects ecosystem functioning: the selection/sampling effect and the complementarity effect (Loreau & Hector 2001). The sampling effect occurs when more diverse communities have higher functioning because they have a higher probability of containing the best-performing species, i.e. the diversity effect is driven by only one (or few) species (Cardinale et al. 2011). In contrast, the complementarity effect occurs when higher ecosystem functioning in more diverse communities is due to niche differences or positive interactions among species, i.e. the diversity effect is driven by many species (Cardinale et al. 2011). Both effects are considered valid and important mechanisms by which diversity affects ecosystem functioning (Cardinale et al. 2006, 2011, 2012; Fargione et al. 2007).

Similarly, the complementarity effect and the sampling effect are important mechanisms by which diversity affects stability (Loreau & Mazancourt 2013; Loreau et al. 2021) and resilience (Steiner et al. 2006; Hughes & Stachowicz 2011). For example, diversity can stabilize aggregate ecosystem properties (e.g. community biomass) when species differ in their responses to environmental factors and therefore respond asynchronously to environmental fluctuations (temporal complementarity effect; Loreau & Mazancourt 2013). However, when considering ecosystem responses to directional environmental change or pulse disturbances rather than to environmental fluctuations, the sampling effect might be of particular importance (Steiner et al. 2006): more diverse systems might absorb a greater amount of change by having a higher chance of containing species with high environmental tolerance. In this scenario, the positive diversity effect on resilience is driven by one (or few) species. From a conservation perspective, however, one would nevertheless strive to maintain diversity and not only the most tolerant species because of the uncertainties associated with environmental change: we do not know with sufficient certainty which stressors will change in a given ecosystem and which traits will therefore be relevant in the future.

@@

For most of the explored combinations and amounts of functional diversity/tolerance the tipping point positions did not differ between the diversity manipulation and the tolerance manipulation (Figure \@ref(fig:max-tol1)). However, in some cases there were effects, three of which are described in the follow sections.

(ref:max-tol1) Effects of strain variation on position of tipping points compared between simulations with all (nine) strains present and simulations where only the most tolerant strain is present.

@@

## remove data where there are no stable states
## this subset is identified by inspection of the figure
#all_stab <- all_stab %>%
#  filter(!(var_treat == "CB-SB" & stand_var > 0.84 & num_strains == -1))

@@

When SB contained multiple strains varying in growth rate (and tolerance) there were only two stable states (Figure \@ref(fig:max-tol2)), as was previously reported. When, however, only one SB strain with very high tolerance and low maximum growth rate was present, there three stable states: the two previously found, and another where all three functional groups coexisted (Figure \@ref(fig:max-tol3)) – this never occurred when there were multiple strains varying in tolerance.

Apparently, at low oxygen diffusivity levels, the absence of SB strains with high growth rate precludes the formation of a stable state that does not include cyanobacteria. It seems that the most tolerant SB strain does not have a high enough growth rate to itself exclude the cyanobacteria. Cyanobacteria are therefore always present, regardless of the oxygen diffusivity. We do not have an explanation of why the cyanobacteria are, however, excluded at intermediate oxygen diffusivities.

@@

```{r max-tol3, echo = FALSE,  include=TRUE, fig.cap='(ref:max-tol3)', fig.height=10, fig.width=8, fig.align="center"}
max_tol_one <- filter(maxtolonly_strains,
       CB_var_gmax_s == 0,
       abs(SB_var_gmax_s - 0.09473684) < 0.001,
       PB_var_gmax_s == 0)
fig_state_vs_o2diff_sidebyside_alternative(max_tol_one)
```

@@

## High tolerance vs. high diversity in CB-SB

In the SB-SB case, presence of only the high tolerance strain could cause the system to not display any alternate stable states (Figure \@ref(fig:max-tol5)).

@@

```{r max-tol5, echo = FALSE,  include=TRUE, fig.cap='(ref:max-tol5)', fig.height=10, fig.width=8, fig.align="center"}
max_tol_one_CBSB <- filter(maxtolonly_strains,
       abs(CB_var_gmax_s - 0.026592798) < 0.001,
       abs(SB_var_gmax_s - 0.079778395) < 0.001,
       PB_var_gmax_s == 0)
max_tol_one_CBSB <- filter(maxtolonly_strains,
       abs(CB_var_gmax_s - 0.028254848) < 0.001,
       abs(SB_var_gmax_s - 0.084764545) < 0.001,
       PB_var_gmax_s == 0)

fig_state_vs_o2diff_sidebyside_alternative(max_tol_one_CBSB)
```

@@

When in each of the three functional groups only the most tolerant strain is present, the resilience of the oxic state is greater than when there is variation in strain tolerance within groups. It is not entirely clear why this effect only occurs only in this combination.

@@

When two functional groups varied in traits, the effects of trait variation were additive when one of the groups had no diversity effect on a tipping point but non-additive when both groups had diversity effects on a tipping point (Fig. \@ref(fig:fig-add-linear) a-f). High trait variation in cyanobacteria and sulfate-reducing bacteria caused the oxic-anoxic tipping point to occur at lower oxygen diffusivity than expected assuming additivity of the individual effects of diversity in these two groups (Fig. \@ref(fig:fig-add-linear)d). When the two groups of sulfur bacteria varied in traits, the anoxic-oxic tipping point occurred at higher oxygen diffusivity than expected by additivity (Fig. \@ref(fig:fig-add-linear)f).

When all three groups varied in traits, only the diversity effect in phototrophic sulfur bacteria on the oxic-anoxic tipping point was additive; all other effects were non-additive, especially at higher trait variation (Fig. \@ref(fig:fig-add-linear)g-i).

# Comparing the effects of different treatments

(ref:comp-comb) Another method for comparing the effects of different diversity treatments. Some lines (treatment combinations) are over-plotted and therefore show identical effects of some combinations.

```{r comp-comb, echo = FALSE,  include=TRUE, fig.height=8, fig.width=8, fig.align="center", fig.cap='(ref:comp-comb)'}

all_stab <- readRDS(here("data",
                         microxanox_version_path,
                         "temporal_method",
                         "event_definition_2",
                         "processed_data",
                         "all_stab_data.RDS"))
wait_time <- 1e6
which_strain <- 9
plot_here <- all_stab %>%
  filter(waittime == wait_time)

  p1 <- plot_here %>%
    #filter(var_treat == "CB") %>%
    filter(num_strains == which_strain) %>%
    filter(Species == "SB_tot") %>%
    ggplot(aes(x = stand_var
               #col = as.factor(num_strains))
    )) +
    geom_line(aes(y = hyst_min_log, col=var_treat), lwd = 1, alpha = 0.8) +
    geom_line(aes(y = hyst_max_log, col=var_treat), lwd = 1, alpha = 0.8) +
    #geom_ribbon(aes(ymin = hyst_min_log, ymax = hyst_max_log),
    #            fill = "green", alpha = 0.1) +
    #facet_grid(var_treat ~ ., scales = "fixed") +
    xlab("Standardised amount of trait variation") +
    ylab(expression(log[10](oxygen~diffusivity)~(h^{-1}))) +
    #labs(fill = "Variation in\nonly these\nfunctional groups") +
    coord_flip() +
    #guides(col = guide_legend(title="Number of strains"),
    #       fill = guide_none()) +
    theme_bw() +
    theme(
      legend.position="top",
      panel.background = element_blank(),
      #strip.text.x = element_blank()
    ) +
    geom_hline(yintercept = c(-8, 0), col = "grey", lwd = 3) #+
    #geom_text(aes(x = 0.95, y = -8.1, label = label), colour = "black") +
    #ggtitle(figure_title)

  p1

# CB diversity effects

Increased CB diversity causes a slight decrease in the resilience of the anoxic state. The following two figures show the state responses to oxygen diffusivity with low Figure \@ref(fig:ss-CB-only1) and high Figure \@ref(fig:ss-CB-only2) levels of CB diversity.

The difference in the position of the anoxic to oxic tipping point is small. It is, in fact the smallest it could be, as it is the smallest step size in oxygen diffusivity used in the simulations. Furthermore, we think the presence of higher strain diversity can increase the rate of response of the system (since there is a strain with higher growth rate than in low diversity levels) and that this difference in response rates is likely responsible for the very small difference in resilience.

(ref:ss-CB-only1) Stable state of the system with **low** level of diversity in only CB, and no diversity in the other two groups. All other aspects are as in Figure \@ref(fig:ss-temp-meth3).

```{r ss-CB-only1, echo = FALSE,  include=TRUE, fig.cap='(ref:ss-CB-only1)', fig.height=10, fig.width=8, fig.align="center"}
wait_time <- "1e+06"
ss_9s <- readRDS(here::here("data",
                            microxanox_version_path, 
                            "temporal_method",
                            "event_definition_2",
                            paste0("ss_data_9strains_waittime", wait_time, "_event_definition_2.RDS")))
#stab_9s <- readRDS(here::here(\"data", "0_ss_finding", "temporal_method", microxanox_version_path, paste0("stab_data_9strains_waittime", wait_time, "_event_definition_2.RDS"))

only_CB_variable <- ss_9s %>%
  filter(
    abs(SB_var_gmax_s) < 0.0001,
    abs(PB_var_gmax_s) < 0.0001
  )
only_CB_variable <- only_CB_variable[order(only_CB_variable$CB_var_gmax_s), ]

#only_CB_variable$CB_var_gmax_s

fig_state_vs_o2diff_sidebyside_alternative(only_CB_variable[2,])

(ref:ss-CB-only2) Stable state of the system with **high** level of diversity in only CB, and no diversity in the other two groups. All other aspects are as in Figure \@ref(fig:ss-temp-meth3).

```{r ss-CB-only2, echo = FALSE,  include=TRUE, fig.cap='(ref:ss-CB-only2)', fig.height=10, fig.width=8, fig.align="center"}

fig_state_vs_o2diff_sidebyside_alternative(only_CB_variable[20,])
